% !TEX root = ../coursera_sc_03.tex

\begin{frame} % название фрагмента

    \videotitle{Black and Scholes model}
    
    \end{frame}
    
    
    \begin{frame}{Black and Scholes model: short plan}
    
      \begin{itemize}[<+->]
        \item \alert{Assumptions} of the model.
        \item The main \alert{question}.
        \item Solving the price \alert{stochastic differential equation}.
      \end{itemize}
    
    \end{frame}


\begin{frame}{BS-model: assumptions \informalduck}
        
\begin{itemize}[<+->]
          \item \alert{Unique} share type is traded. The share price $S_t$ satisfies 
          \[
          dS_t = \mu S_t \, dt + \sigma S_t  \, dW_t.
          \]
          \item The \alert{risk free} interest rate $r$ is constant and unique for all horizons. 
          \[
          dB_t = r B_t \, dt  \quad B_t = B_0 \exp(rt), \; B_0 =1.
          \]
          \item No taxes, no dividends, \alert{no transaction costs}. 
          \item Time is continuous, shares are \alert{infinitely divisible}.
          \item \alert{Short selling} is allowed.  
          \item \alert{No arbitrage} opportunities.
          \item The parameters $r$, $\sigma$, $\mu$ \alert{are known}.
\end{itemize}
        
\end{frame}

\begin{frame}{BS-model}

\begin{block}{The Question}
  How can we calculate the price of a particular financial claim?  
\end{block}

\pause

\alert{Example}. At time moment $T=2$ John would like to get $10$ if the $S_2 > 100$ and nothing otherwise. 

\pause 

What is the fair price John should pay at $t=0$?

\end{frame}


\begin{frame}
  \frametitle{SDE for the price}
  
  \alert{Stochastic differential equation} for the price:
  \[
  dS_t = \mu S_t \, dt + \sigma S_t \, dW_t.
  \]

  \pause

  In \alert{full} form:
  \[
  S_t = S_0 + \int_0^t \mu S_u \, du + \int_0^t \sigma S_u \, dW_u.  
  \]

\end{frame}

\begin{frame}
  \frametitle{Solving the SDE \knightduck}
  \alert{Stochastic differential equation} for the price:
  \[
  dS_t = \mu S_t \, dt + \sigma S_t \, dW_t.
  \]
  
  \onslide<2-> {By \alert{mnemonic} rules, }
  \begin{flalign*}
    \onslide<3-> {(dS_t)^2 = (\mu S_t \, dt + \sigma S_t \, dW_t)^2 =}\onslide<4->{ 0 + 0 + \sigma^2 S_t^2 dt.}
  \end{flalign*} 

  \onslide<5->{Let's consider $Y_t = \ln S_t$. Accordign to \alert{Itô's lemma},}
\begin{flalign*}
    \onslide<6->{dY_t & = }\onslide<7->{\frac{1}{S_t} \, dS_t + 0 \, dt + }\onslide<8->{ \frac{1}{2}\left(\frac{-1}{S_t^2}\right) (dS_t)^2 =&}\\
    \onslide<9->{=\mu \, dt + \sigma dW_t - \frac{1}{2}\sigma^2 \, dt.&}
\end{flalign*} 
  
\end{frame}


\begin{frame}
  \frametitle{Solving the SDE\ldots}

  The log-price $Y_t = \ln S_t$ in \alert{short} form,
  \[
  dY_t = \left( \mu - \frac{\sigma^2}{2}\right)\, dt + \sigma \, dW_t.  
  \]
\pause
  In \alert{full} form,
  \[
    Y_t = Y_0 + \int_0^t\left( \mu - \frac{\sigma^2}{2}\right)\, du + \int_0^t \sigma \, dW_u
  \]
\pause
  Without integrals, 
  \[
  \ln S_t = \ln S_0 +   \left( \mu - \frac{\sigma^2}{2}\right)t + \sigma W_t.
  \]
\pause
  \alert{Finally}, 
  \[
  S_t = S_0 \exp  \left(  \left( \mu - \frac{\sigma^2}{2}\right)t + \sigma W_t \right).
  \]

\end{frame}

\begin{frame}{Itô's lemma: summary}
  
  \begin{itemize}[<+->]
      \item With BS model we will calculate the \alert{fair} price.
      \item The share price follows \alert{geometric} brownian motion, 
      \[
        S_t = S_0 \exp  \left(  \left( \mu - \frac{\sigma^2}{2}\right)t + \sigma W_t \right).
      \]

      \item \alert{Risk free rate} $r$ is constant.
      \item There are \alert{no arbitrage} opportunities.
  \end{itemize}
    
\end{frame}
  